### 2024-01-01 chapters/paper_hierarchical.tex
Update top reference
Change image to png
Fix commas after equations

\textit{This chapter is a piece of original research published as part of the project:} \newline
\begin{center}
    \begin{enumerate}[leftmargin=8mm,rightmargin=8mm,topsep=0mm,label={[\Alph*]}]
        \setcounter{enumi}{0}
        \item \fullcite{havtorn_hierarchical_2021} \main \. \parencite{havtorn_hierarchical_2021}
        \end{enumerate}
\end{center}

@@

\begin{figure}[t]
    \centering
    %\vspace{0.17cm}
    \includegraphics[width=1\columnwidth]{paper_hierarchical/biva_reconstructions.pdf}
    % https://docs.google.com/drawings/d/1UXdA3c18Oek9_env1vEB9mqOAHZ-tu0ijWmvqmapWCk/edit?usp=sharing
    \vspace{0mm}
    \caption[Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables.]{Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables  \parencite{maaloe_biva_2019}.
    The higher the latent variable, the more the reconstructions fall into the mode of the learned distribution.
    It is more common to wear regular glasses than sunglasses but most common not to wear glasses at all.
    A man with long hair collapses into the mode of the more common long-haired woman.}
    \vspace{0mm}
    \label{fig_hierarchical:biva-reconstructions-celeba}
\end{figure}
Finally, we look at reconstructions of out-of-distribution data.
\cref{fig_hierarchical:reconstructions-fashionmnist} illustrates that MNIST data is surprisingly well reconstructed by a hierarchical VAE trained on FashionMNIST.
Similar results have been found elsewhere \parencite{xiao_likelihood_2020}.
We repeat the previous experiment and replace inference distributions by their corresponding conditional prior, and now observe that reconstructions from higher latent layers become increasingly similar to the data on which the model was trained.
The reliance on conditional priors seems to prevent accurate reconstruction of out-of-distribution data.
Some details are lost on in-distribution data too, but the distinction between that and out-of-distribution data becomes more clear.

@@

\subsection{Variational autoencoders}
The variational autoencoder (VAE) \parencite{kingma_autoencoding_2014, rezende_stochastic_2014} is a framework for constructing deep generative models defined by an observed variable $\mathbf{x}$ and a stochastic latent variable $\mathbf{z}$.
Typically, a neural network with parameters $\thetab$ is chosen to parameterize the generative distribution $p_\thetab(\xb,\zb)=p_\thetab(\xb|\zb)p(\zb)$, where the prior $p(\zb)$ is commonly a standard Gaussian $\mathcal{N}(\0, \Ib)$.
The true posterior $p(\zb|\xb)$ is generally not analytically tractable and is approximated by a variational distribution $q_\phib(\zb|\xb)$ parameterized via another neural network with parameters $\phib$. The approximate posterior $q_\phib(\zb|\xb)$ is most often  a diagonal covariance Gaussian.
The model parameters $\thetab$ and variational parameters $\phib$ are jointly optimized by maximizing the \textit{evidence lower bound} (ELBO),
\begin{equation}\label{eq_hierarchical:elbo}
    \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}(\xb; \thetab, \phib)\ .
\end{equation}
For brevity, we will denote $\mathcal{L}(\xb; \thetab, \phib)$ as $\mathcal{L}(\xb)$ or $\mathcal{L}$. The reparameterization trick is used to backpropagate gradients through the stochastic latent variables with low variance.

@@

\subsection{Hierarchical variational autoencoders}\label{sec_paper_hierarchical:background-hie-VAE}
Hierarchical VAEs are a family of probabilistic latent variable models which extends the basic VAE by introducing a hierarchy of $L$ latent variables $\zb=\zb^{(1)}, \dots, \zb^{(L)}$.
The most common generative model is defined from the top down as $p_\thetab(\xb|\zb)=p_\thetab(\xb|\zb^{(1)})p_\thetab(\zb^{(1)}|\zb^{(2)})\cdots p_\thetab(\zb^{(L-1)}|\zb^{(L)})$.
The inference model can then be defined in two ways respectively referred to as \textit{bottom-up} \parencite{burda_importance_2016}
\begin{equation}
    %q_\phib(\zb|\xb)=q_\phib(\zb^{(1)}|\xb)q_\phib(\zb^{(2)}|\zb^{(1)})\cdots q_\phib(\zb^{(L)}|\zb^{(L-1)})
    q_\phib(\zb|\xb) = q_\phib(\zb^{(1)}|\xb)\textstyle\prod_{i=2}^{L} q_\phib(\zb^{(i)}|\zb^{(i-1)})
\end{equation}
and \textit{top-down} \parencite{sonderby_ladder_2016}
\begin{equation}
    % q_\phib(\zb|\xb)=q_\phib(\zb^{(L)}|\xb)q_\phib(\zb^{(L-1)}|\zb^{(L)})\cdots q_\phib(\zb^{(2)}|\zb^{(1)}),
    q_\phib(\zb|\xb) = q_\phib(\zb^{(L)}|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb^{(i)}|\zb^{(i+1)}) \ .
\end{equation}
Regardless of the choice of inference model, a hierarchical VAE is still trained using the ELBO \cref{eq_hierarchical:elbo}.

@@

\section{OOD detection with hierarchical VAEs}
\subsection{A bound for semantic OOD detection}
If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \crefrange{fig_hierarchical:reconstructions-fashionmnist}{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb^{(1)})$ will be high for both in- and out-of-distribution data.
Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb^{(1)})] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection. For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel p(\zb))$ can be expressed by a hierarchical sum,% over the hierarchy
% \begin{multline}
%     \mathcal{L}(\xb) =  \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \log p_\thetab(\xb|\zb^{(1)}) \\
%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} 
%                      + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] .
% \end{multline}
% \begin{multline}
%     D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel p(\zb)) = \\
%     \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] .
% \end{multline}
\begin{equation}
    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] \ .
\end{equation}
In general, the absolute log-ratios grow with $\mathrm{dim}(\zb^{(i)})$ as the individual log probability terms are computed by summing over the dimensionality of $\zb^{(i)}$.
This means that the value of the KL-term is dominated by terms where $\zb^{(i)}$ is high-dimensional. We refer to \cref{sec_hierarchical:analysis} for a more detailed argument.
Since hierarchical VAEs are generally constructed with a bottleneck type structure, the terms corresponding to latent variables towards the top of the hierarchy will have a vanishing influence on the value of the KL-term.
However, as the semantic information most relevant for OOD detection has a tendency to be represented in the top-most latent variables, this makes OOD detection using the regular ELBO difficult, even for state-of-the-art models.
This behavior has also been reported by \textcite{xiao_likelihood_2020}.

@@

We can gather further insights about this score if we write the regular ELBO and the $\mathcal{L}^{>k}$ bounds in the exact form that includes the intractable KL-divergence between the approximate and true posteriors,
\begin{align}
    \mathcal{L}      &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) \parallel p_\thetab(\zb|\xb)\right), \label{eq_hierarchical:likelihoods-as-exact} \\ 
    \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb)\right) \nonumber \ .
\end{align}
Subtracting these cancel out the two data likelihood terms $\log p_\thetab(\xb)$ and only the KL-divergences from the approximate to the true posterior remain,
\begin{align}\label{eq_hierarchical:llr-as-kls}
    LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) \parallel p_\thetab(\zb|\xb)\right) \\
                 &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb)\right) \ . \notag
\end{align}

>>>>

\textit{This chapter is a piece of original research by the candidate, previously published elsewhere and reprinted here with permission:} \newline
\begin{center}
    \begin{enumerate}[leftmargin=8mm,rightmargin=8mm,topsep=0mm,label={[\Alph*]}]
        \setcounter{enumi}{0}
        \item \fullcite{havtorn_hierarchical_2021} \main \hspace{0.1em} \parencite{havtorn_hierarchical_2021}
        \end{enumerate}
\end{center}

@@

\begin{figure}[t]
    \centering
    %\vspace{0.17cm}
    \includegraphics[width=1\columnwidth]{paper_hierarchical/biva_reconstructions.png}
    % https://docs.google.com/drawings/d/1UXdA3c18Oek9_env1vEB9mqOAHZ-tu0ijWmvqmapWCk/edit?usp=sharing
    \vspace{0mm}
    \caption[Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables.]{Reconstructions of in-distribution data (CelebA) of the BIVA model using higher latent variables  \parencite{maaloe_biva_2019}.
    The higher the latent variable, the more the reconstructions fall into the mode of the learned distribution.
    It is more common to wear regular glasses than sunglasses but most common not to wear glasses at all.
    A man with long hair collapses into the mode of the more common long-haired woman.}
    \vspace{0mm}
    \label{fig_hierarchical:biva-reconstructions-celeba}
\end{figure}
Finally, we look at reconstructions of out-of-distribution data.
\cref{fig_hierarchical:reconstructions-fashionmnist} illustrates that MNIST data is surprisingly well reconstructed by a hierarchical VAE trained on FashionMNIST.
Similar results have been found elsewhere \parencite{xiao_likelihood_2020}.
We repeat the previous experiment and replace inference distributions by their corresponding conditional prior, and now observe that reconstructions from higher latent layers become increasingly similar to the data on which the model was trained.
The reliance on conditional priors seems to prevent accurate reconstruction of out-of-distribution data.
Some details are lost on in-distribution data too, but the distinction between that and out-of-distribution data becomes more clear.

@@

\subsection{Variational autoencoders}
The variational autoencoder (VAE) \parencite{kingma_autoencoding_2014, rezende_stochastic_2014} is a framework for constructing deep generative models defined by an observed variable $\mathbf{x}$ and a stochastic latent variable $\mathbf{z}$.
Typically, a neural network with parameters $\thetab$ is chosen to parameterize the generative distribution $p_\thetab(\xb,\zb)=p_\thetab(\xb|\zb)p(\zb)$, where the prior $p(\zb)$ is commonly a standard Gaussian $\mathcal{N}(\0, \Ib)$.
The true posterior $p(\zb|\xb)$ is generally not analytically tractable and is approximated by a variational distribution $q_\phib(\zb|\xb)$ parameterized via another neural network with parameters $\phib$. The approximate posterior $q_\phib(\zb|\xb)$ is most often  a diagonal covariance Gaussian.
The model parameters $\thetab$ and variational parameters $\phib$ are jointly optimized by maximizing the \textit{evidence lower bound} (ELBO),
\begin{equation}\label{eq_hierarchical:elbo}
    \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}(\xb; \thetab, \phib) \enspace .
\end{equation}
For brevity, we will denote $\mathcal{L}(\xb; \thetab, \phib)$ as $\mathcal{L}(\xb)$ or $\mathcal{L}$. The reparameterization trick is used to backpropagate gradients through the stochastic latent variables with low variance.

@@

\subsection{Hierarchical variational autoencoders}\label{sec_paper_hierarchical:background-hie-VAE}
Hierarchical VAEs are a family of probabilistic latent variable models which extends the basic VAE by introducing a hierarchy of $L$ latent variables $\zb=\zb^{(1)}, \dots, \zb^{(L)}$.
The most common generative model is defined from the top down as $p_\thetab(\xb|\zb)=p_\thetab(\xb|\zb^{(1)})p_\thetab(\zb^{(1)}|\zb^{(2)})\cdots p_\thetab(\zb^{(L-1)}|\zb^{(L)})$.
The inference model can then be defined in two ways respectively referred to as \textit{bottom-up} \parencite{burda_importance_2016}
\begin{equation}
    %q_\phib(\zb|\xb)=q_\phib(\zb^{(1)}|\xb)q_\phib(\zb^{(2)}|\zb^{(1)})\cdots q_\phib(\zb^{(L)}|\zb^{(L-1)})
    q_\phib(\zb|\xb) = q_\phib(\zb^{(1)}|\xb)\textstyle\prod_{i=2}^{L} q_\phib(\zb^{(i)}|\zb^{(i-1)})
\end{equation}
and \textit{top-down} \parencite{sonderby_ladder_2016}
\begin{equation}
    % q_\phib(\zb|\xb)=q_\phib(\zb^{(L)}|\xb)q_\phib(\zb^{(L-1)}|\zb^{(L)})\cdots q_\phib(\zb^{(2)}|\zb^{(1)}),
    q_\phib(\zb|\xb) = q_\phib(\zb^{(L)}|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb^{(i)}|\zb^{(i+1)}) \enspace .
\end{equation}
Regardless of the choice of inference model, a hierarchical VAE is still trained using the ELBO \cref{eq_hierarchical:elbo}.

@@

\section{OOD detection with hierarchical VAEs}
\subsection{A bound for semantic OOD detection}
If the lowest latent variable in the VAE hierarchy codes for a large part of the low-level features required to reconstruct the input with high accuracy, as exemplified in \crefrange{fig_hierarchical:reconstructions-fashionmnist}{fig_hierarchical:biva-reconstructions-celeba}, then $p_\thetab(\xb|\zb^{(1)})$ will be high for both in- and out-of-distribution data.
Hence, any OOD detection capabilities based on the ELBO $\mathcal{L} = \mathbb{E}_{q_\phib(\zb|\xb)}[\log p_\thetab(\xb|\zb^{(1)})] - D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel  p(\zb))$ from \cref{eq_hierarchical:elbo} relies on the KL-term for OOD detection. For a bottom-up hierarchical VAE, the KL-term $D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel p(\zb))$ can be expressed by a hierarchical sum,% over the hierarchy
% \begin{multline}
%     \mathcal{L}(\xb) =  \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \log p_\thetab(\xb|\zb^{(1)}) \\
%                      + \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} 
%                      + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] .
% \end{multline}
% \begin{multline}
%     D_{\mathrm{KL}}( q_\phib(\zb|\xb) \parallel p(\zb)) = \\
%     \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] .
% \end{multline}
\begin{equation}
    \mathbb{E}_{q_\phib(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\thetab(\zb^{(i)}|\zb^{(i+1)})}{q_\phib(\zb^{(i)}|\zb^{(i-1)})} + \log \frac{p_\thetab(\zb^{(L)})}{q_\phib(\zb^{(L)}|\zb^{(L-1)})} \Big] \enspace .
\end{equation}
In general, the absolute log-ratios grow with $\mathrm{dim}(\zb^{(i)})$ as the individual log probability terms are computed by summing over the dimensionality of $\zb^{(i)}$.
This means that the value of the KL-term is dominated by terms where $\zb^{(i)}$ is high-dimensional. We refer to \cref{sec_hierarchical:analysis} for a more detailed argument.
Since hierarchical VAEs are generally constructed with a bottleneck type structure, the terms corresponding to latent variables towards the top of the hierarchy will have a vanishing influence on the value of the KL-term.
However, as the semantic information most relevant for OOD detection has a tendency to be represented in the top-most latent variables, this makes OOD detection using the regular ELBO difficult, even for state-of-the-art models.
This behavior has also been reported by \textcite{xiao_likelihood_2020}.

@@

We can gather further insights about this score if we write the regular ELBO and the $\mathcal{L}^{>k}$ bounds in the exact form that includes the intractable KL-divergence between the approximate and true posteriors,
\begin{align}
    \mathcal{L}      &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) \parallel p_\thetab(\zb|\xb)\right), \label{eq_hierarchical:likelihoods-as-exact} \\ 
    \mathcal{L}^{>k} &= \log p_\thetab(\xb) - D_{\mathrm{KL}}\left( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb)\right) \nonumber \enspace .
\end{align}
Subtracting these cancel out the two data likelihood terms $\log p_\thetab(\xb)$ and only the KL-divergences from the approximate to the true posterior remain,
\begin{align}\label{eq_hierarchical:llr-as-kls}
    LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phib(\zb|\xb) \parallel p_\thetab(\zb|\xb)\right) \\
                 &\quad + D_{\mathrm{KL}}\left( p_\thetab(\zb^{(\leq k)}|\zb^{(>k)}) q_\phib(\zb^{(>k)}|\xb) \parallel p_\thetab(\zb|\xb)\right) \enspace . \notag
\end{align}

### 2024-01-01 chapters/paper_modelagnostic.tex
Fix references
Add citation to top reference
Update initial reference
Update top reference
Fix commas after equations

\textit{This chapter is a piece of original research published as part of the project:} \newline
\begin{center}
    \begin{enumerate}[leftmargin=8mm,rightmargin=8mm,topsep=0mm,label={[\Alph*]}]
        \setcounter{enumi}{1}
        \item \fullcite{bergamin_modelagnostic_2022} \co \label{enumerate: paper-modelagnostic}
        \end{enumerate}
\end{center}

@@

More specifically, given a kernel whose feature map is $\Phi : \mathcal{X} \to \mathcal{H}$, the MMD between two distributions $P$ and $Q$ over $\mathcal{X}$ is defined as
\begin{equation}
    \textup{MMD}_\Phi(P,Q) = \| E_{X\sim P}[\Phi(X)] - E_{Y\sim Q}[\Phi(Y)] \|_\mathcal{H}. \hspace{-.1em}
\end{equation}
In our context, the test statistics will be of the form
\begin{equation}
    \label{eq_modelagnostic:mmd}
    \textup{MMD}_\Phi\left(\frac{1}{m}\sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \left\|\frac{1}{m}\sum_{i=1}^{m} \Phi(x_i) -  \frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)\right\|_\mathcal{H},
\end{equation}
where $\Phi$ is a kernel feature map built using the generative model and $x_1,\ldots,x_m$ is the training data, i.e.\@ samples from  $p_\textup{data}$. When $\mathcal{H}$ is a simple finite-dimensional Hilbert space and $\Phi$ can be computed easily, then \cref{eq_modelagnostic:mmd} can be computed by going through the data and computing the means in an online fashion.% While the definition seems to imply that we can only perform group-sample OOD detection, it is straightforward to notice that any test statistic of the form of \cref{eq_modelagnostic:mmd} can be used also for single-test OOD detection when $n=1$.%by just substituting  $\frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)$ with simply $\Phi(\tilde{x})$.

@@

\paragraph{The Fisher kernel} An important example of kernel based on a generative model is the \emph{Fisher kernel} of \textcite{jaakkola_exploiting_1999}. The embedding of this kernel is the Fisher score 
\begin{equation}
    \label{eq_modelagnostic:Fisher_score}
    \Phi_\textup{Fisher} (x) = I(\theta)^{-\frac{1}{2}} \nabla \log p_\theta (x),
\end{equation}

and the corresponding reproducing kernel Hilbert space norm is just the $\ell_2$ norm: $|| \cdot || _\mathcal{H} = || \cdot ||_2$. In the case of the Fisher kernel, this means that \cref{eq_modelagnostic:mmd} becomes:
\begin{multline}
    \label{eq_modelagnostic:mmd_fisher}
    \textup{MMD}_{\Phi_\textup{Fisher}}\left(\frac{1}{m} \sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\ \left\|\frac{I(\theta)^{-\frac{1}{2}}}{m} \sum_{i=1}^{m}  \nabla \log p_\theta (x_i) - \frac{I(\theta)^{-\frac{1}{2}}}{n}  \sum_{i=1}^{n}  \nabla \log p_\theta (\tilde{x}_i)\right\|_2.
\end{multline}
We will see later that MMD with a Fisher kernel is closely related to the score statistic. In \cref{appendix_modelagnostic:mahalanobis}, we additionally show that another popular OOD metric known as the \emph{Mahalanobis score} \parencite{lee_simple_2018} can be interpreted as a MMD statistic with a certain Fisher kernel.

\paragraph{The typicality kernel} A very simple approach of embedding the data using $p_\theta$ is to choose $\Phi_\textup{Typical}(x) = \log p_\theta (x)$. Then, MMD is exactly equivalent to the \emph{typicality test statistic} of \textcite{nalisnick_detecting_2019}, although this connection was not explicitly stated  by \textcite{nalisnick_detecting_2019}. Because of this, we call the kernel $k(x,y) = \log p_\theta (x) \cdot \log p_\theta (y)$ the \emph{typicality kernel}.  %We could also call this "new" kernel the "likelihood kernel".
%While $\Phi_\textup{Typical}$ is not as well motivated as the Fisher kernel, we found that it generally gives good results. 
While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviours of probability distributions in high-dimensional space as highlighted by \textcite{nalisnick_deep_2019}. We also found that using this kernel generally gives good results for OOD tasks. An interesting analysis that we'd not consider in this paper would be to study the properties of this kernel.

@@

\paragraph{Combining test statistics by combining $p$-values}
Fisher's \citeyearpar{fisher_statistical_1925} method is a procedure to combine different $p$-values $p_1,\dots, p_k$. This method assumes that all the considered test statistics are independent, and \textcite{folks_asymptotic_1971} proved that it is asymptotically optimal among all methods of combining independent tests.
%
Given $T_1,\dots, T_k$ and corresponding $p$-values $p_1,\dots, p_k$, Fisher's method combines the $p$-values into a test statistic $X^2$ defined as
\begin{equation}
    \label{eq_modelagnostic:Fisher_method}
    X^2 \sim -2 \sum_{j=1}^{k} \ln (p_j).
\end{equation}
In case all null-hypotheses are accepted, the resulting test statistic $X^2$ follows a chi-squared distribution with $2k$ degrees of freedom. 
%
In the \cref{appendix_modelagnostic:harmonic}, we also consider the Harmonic mean $p$-value \parencite{wilson_harmonic_2019} as a way to combine $p$-values from different statistics. This method usually works best when the statistics are not independent.

@@

Assume that we have two independent OOD data samples that follow a product of truncated normal distributions, with density proportional to $$\mathcal{N}(x | 0,I_D) \cdot  \mathbf{1}\{x_1>0,\ldots,x_D>0\}.$$

We denote by $T_\text{score}^\text{ood},T_\text{score}^\text{id}$ and $T_\text{typicality}^\text{ood},T_\text{typicality}^\text{id}$ the statistics obtained when confronted with either OOD data from the truncated normal, or the in-distribution data. While these statistics are random in general, they will become deterministic when $D \to \infty$, by virtue of the law of large numbers.

For the typicality statistic, these two OOD samples will be indistinguishable from Gaussian ones. Indeed, when $D \to \infty$, both $T_\text{typicality}^\text{ood}$ and $T_\text{typicality}^\text{id}$ will be $\mathcal{O}(D)$.
%
On the other hand, for the score, one can show that
\begin{equation}
T_\text{score}^\text{ood} - T_\text{score}^\text{id} \sim 2 D \mu_\text{TN}^2,
\end{equation}
where $\mu_\text{TN}>0$ is the mean of the truncated normal distribution.

>>>>

\textit{This chapter is a piece of original research by the candidate, previously published elsewhere and reprinted here with permission:} \newline
\begin{center}
    \begin{enumerate}[leftmargin=8mm,rightmargin=8mm,topsep=0mm,label={[\Alph*]}]
        \setcounter{enumi}{1}
        \item \fullcite{bergamin_modelagnostic_2022} \co \hspace{0.1em} \parencite{bergamin_modelagnostic_2022}
        \end{enumerate}
\end{center}

@@

More specifically, given a kernel whose feature map is $\Phi : \mathcal{X} \to \mathcal{H}$, the MMD between two distributions $P$ and $Q$ over $\mathcal{X}$ is defined as
\begin{equation}
    \textup{MMD}_\Phi(P,Q) = \| E_{X\sim P}[\Phi(X)] - E_{Y\sim Q}[\Phi(Y)] \|_\mathcal{H} \enspace . \hspace{-.1em}
\end{equation}
In our context, the test statistics will be of the form
\begin{equation}
    \label{eq_modelagnostic:mmd}
    \textup{MMD}_\Phi\left(\frac{1}{m}\sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \left\|\frac{1}{m}\sum_{i=1}^{m} \Phi(x_i) -  \frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)\right\|_\mathcal{H} \enspace ,
\end{equation}
where $\Phi$ is a kernel feature map built using the generative model and $x_1,\ldots,x_m$ is the training data, i.e.\@ samples from  $p_\textup{data}$. When $\mathcal{H}$ is a simple finite-dimensional Hilbert space and $\Phi$ can be computed easily, then \cref{eq_modelagnostic:mmd} can be computed by going through the data and computing the means in an online fashion.% While the definition seems to imply that we can only perform group-sample OOD detection, it is straightforward to notice that any test statistic of the form of \cref{eq_modelagnostic:mmd} can be used also for single-test OOD detection when $n=1$.%by just substituting  $\frac{1}{n}\sum_{i=1}^{n} \Phi(\tilde{x}_i)$ with simply $\Phi(\tilde{x})$.

@@

\paragraph{The Fisher kernel} An important example of kernel based on a generative model is the \emph{Fisher kernel} of \textcite{jaakkola_exploiting_1999}. The embedding of this kernel is the Fisher score 
\begin{equation}
    \label{eq_modelagnostic:Fisher_score}
    \Phi_\textup{Fisher} (x) = I(\theta)^{-\frac{1}{2}} \nabla \log p_\theta (x) \enspace ,
\end{equation}

and the corresponding reproducing kernel Hilbert space norm is just the $\ell_2$ norm: $|| \cdot || _\mathcal{H} = || \cdot ||_2$. In the case of the Fisher kernel, this means that \cref{eq_modelagnostic:mmd} becomes:
\begin{multline}
    \label{eq_modelagnostic:mmd_fisher}
    \textup{MMD}_{\Phi_\textup{Fisher}}\left(\frac{1}{m} \sum_{i=1}^{m} x_i,\frac{1}{n}\sum_{i=1}^n \tilde{x}_i\right) = \\ \left\|\frac{I(\theta)^{-\frac{1}{2}}}{m} \sum_{i=1}^{m}  \nabla \log p_\theta (x_i) - \frac{I(\theta)^{-\frac{1}{2}}}{n}  \sum_{i=1}^{n}  \nabla \log p_\theta (\tilde{x}_i)\right\|_2 \enspace .
\end{multline}
We will see later that MMD with a Fisher kernel is closely related to the score statistic. In \cref{appendix_modelagnostic:mahalanobis}, we additionally show that another popular OOD metric known as the \emph{Mahalanobis score} \parencite{lee_simple_2018} can be interpreted as a MMD statistic with a certain Fisher kernel.

\paragraph{The typicality kernel} A very simple approach of embedding the data using $p_\theta$ is to choose $\Phi_\textup{Typical}(x) = \log p_\theta (x)$. Then, MMD is exactly equivalent to the \emph{typicality test statistic} of \textcite{nalisnick_detecting_2019}, although this connection was not explicitly stated  by \textcite{nalisnick_detecting_2019}. Because of this, we call the kernel $k(x,y) = \log p_\theta (x) \cdot \log p_\theta (y)$ the \emph{typicality kernel}.  %We could also call this "new" kernel the "likelihood kernel".
%While $\Phi_\textup{Typical}$ is not as well motivated as the Fisher kernel, we found that it generally gives good results. 
While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviors of probability distributions in high-dimensional space as highlighted by \textcite{nalisnick_deep_2019}. We also found that using this kernel generally gives good results for OOD tasks. An interesting analysis that we'd not consider in this paper would be to study the properties of this kernel.

@@

\paragraph{Combining test statistics by combining $p$-values}
Fisher's \citeyearpar{fisher_statistical_1925} method is a procedure to combine different $p$-values $p_1,\dots, p_k$. This method assumes that all the considered test statistics are independent, and \textcite{folks_asymptotic_1971} proved that it is asymptotically optimal among all methods of combining independent tests.
%
Given $T_1,\dots, T_k$ and corresponding $p$-values $p_1,\dots, p_k$, Fisher's method combines the $p$-values into a test statistic $X^2$ defined as
%
\begin{equation}
    \label{eq_modelagnostic:Fisher_method}
    X^2 \sim -2 \sum_{j=1}^{k} \ln (p_j) \enspace .
\end{equation}
%
In case all null-hypotheses are accepted, the resulting test statistic $X^2$ follows a chi-squared distribution with $2k$ degrees of freedom. 
%
In the \cref{appendix_modelagnostic:harmonic}, we also consider the Harmonic mean $p$-value \parencite{wilson_harmonic_2019} as a way to combine $p$-values from different statistics. This method usually works best when the statistics are not independent.

@@

Assume that we have two independent OOD data samples that follow a product of truncated normal distributions, with density proportional to
%
\begin{equation}
    \mathcal{N}(x | 0,I_D) \cdot  \mathbf{1}\{x_1>0,\ldots,x_D>0\} \enspace .
\end{equation}
%
We denote by $T_\text{score}^\text{ood},T_\text{score}^\text{id}$ and $T_\text{typicality}^\text{ood},T_\text{typicality}^\text{id}$ the statistics obtained when confronted with either OOD data from the truncated normal, or the in-distribution data. While these statistics are random in general, they will become deterministic when $D \to \infty$, by virtue of the law of large numbers.

For the typicality statistic, these two OOD samples will be indistinguishable from Gaussian ones. Indeed, when $D \to \infty$, both $T_\text{typicality}^\text{ood}$ and $T_\text{typicality}^\text{id}$ will be $\mathcal{O}(D)$.
%
On the other hand, for the score, one can show that
\begin{equation}
T_\text{score}^\text{ood} - T_\text{score}^\text{id} \sim 2 D \mu_\text{TN}^2 \enspace ,
\end{equation}
where $\mu_\text{TN}>0$ is the mean of the truncated normal distribution.
